Neural correlates of the deployment of spatial attention, and their modulation by repetitive movements

### 2024-09-16 manuscript.qmd
Fixed typo causing problem with confidence intervals

if (processdata<2){
  # download processed data if missing
  if (!file.exists('local/Expt1resps.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Expt1resps.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
  if (!file.exists('local/Expt2resps.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Expt2resps.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
  if (!file.exists('local/Exp1Group.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Exp1Group.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
    if (!file.exists('local/Exp2Group.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Exp2Group.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
}

@@

  CIdecoding <- decodingdata
  CIdecoding$Cue <- apply(allCue,1,sd)/sqrt(ncol(allCue))*1.96
  CIdecoding$TargetU <- apply(allTargetU,1,sd)/sqrt(ncol(allTargetU))*1.96
  CIdecoding$CongIncong <- apply(allCongIncong,1,sd)/sqrt(ncol(allCongIncong))*1.96
  CIdecoding$TargetB <- apply(allTargetB,1,sd)/sqrt(ncol(allTargetB))*1.96
  CIdecoding$CueNoStim <- apply(allCueNostim,1,sd)/sqrt(ncol(allCueNostim))*1.96
  CIdecoding$CueStim <- apply(allCueStim,1,sd)/sqrt(ncol(allCueStim))*1.96

@@

  CIdecoding <- decodingdata
  CIdecoding$Cue <- apply(allCue[,stimindices],1,sd)/sqrt(ncol(allCue[,stimindices]))*1.96
  CIdecoding$TargetU <- apply(allTargetU[,stimindices],1,sd)/sqrt(ncol(allTargetU[,stimindices]))*1.96
  CIdecoding$CongIncong <- apply(allCongIncong[,stimindices],1,sd)/sqrt(ncol(allCongIncong[,stimindices]))*1.96
  CIdecoding$TargetB <- apply(allTargetB[,stimindices],1,sd)/sqrt(ncol(allTargetB[,stimindices]))*1.96
  CIdecoding$CueNoStim <- apply(allCueNostim[,stimindices],1,sd)/sqrt(ncol(allCueNostim[,stimindices]))*1.96
  CIdecoding$CueStim <- apply(allCueStim[,stimindices],1,sd)/sqrt(ncol(allCueStim[,stimindices]))*1.96

@@

        decodingBFs2 <- matrix(0,nrow=1201,ncol=7)
  for (t in 1:nrow(allCue)){
    decodingBFs2[t,1] <- extractBF(ttestBF(allCue[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,2] <- extractBF(ttestBF(allTargetU[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,3] <- extractBF(ttestBF(allCongIncong[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,4] <- extractBF(ttestBF(allTargetB[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,5] <- extractBF(ttestBF(allCueNostim[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,6] <- extractBF(ttestBF(allCueStim[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,7] <- extractBF(ttestBF(allCueStim[t,nostimindices],allCueNostim[t,nostimindices],paired=TRUE))$bf
  }
  CIdecoding <- decodingdata
  CIdecoding$Cue <- apply(allCue[,nostimindices],1,sd)/sqrt(ncol(allCue[,nostimindices]))*1.96
  CIdecoding$TargetU <- apply(allTargetU[,nostimindices],1,sd)/sqrt(ncol(allTargetU[,nostimindices]))*1.96
  CIdecoding$CongIncong <- apply(allCongIncong[,nostimindices],1,sd)/sqrt(ncol(allCongIncong[,nostimindices]))*1.96
  CIdecoding$TargetB <- apply(allTargetB[,nostimindices],1,sd)/sqrt(ncol(allTargetB[,nostimindices]))*1.96
  CIdecoding$CueNoStim <- apply(allCueNostim[,nostimindices],1,sd)/sqrt(ncol(allCueNostim[,nostimindices]))*1.96
  CIdecoding$CueStim <- apply(allCueStim[,nostimindices],1,sd)/sqrt(ncol(allCueStim[,nostimindices]))*1.96

@@

A combined EEG and TMS study by @Capotosto2012 demonstrated that the application of TMS to the right intraparietal sulcus (IPS) impairs target detection, particularly on incongruent trials. They additionally noted that the P3 response amplitude was affected by this; on incongruent trials its amplitude was significantly lower, and on congruent trials it was significantly higher. The IPS has been previously identified as an important region in attentional deployment by @Vossel2009 in an fMRI study identifying regions that are selectively active on congruent and incongruent trials. The right IPS and, additionally, the right inferior frontal gyrus were implicated in attentional processes related to both types of trial. @Peelen2004 concluded that both endogenous and exogenous orienting cues involve the same network in their fMRI study, implicating frontal and parietal regions (the premotor cortex, posterior parietal cortex, medial frontal cortex, and right inferior frontal cortex). @Fitzgerald2015 conducted an MRI study using the Posner cueing task and found that activation in the ventral attention network differed between individuals with autism and controls; individuals with autism also showed weaker functional connectivity in the dorsal attentional network compared to their neurotypical counterparts.

@@

This study involved two experiments. In the first experiment, participants completed a Posner cueing task while EEG recordings were taken; a pattern classifier then attempted to differentiate, using the participants’ brain activity, whether they attended to the left vs. to the right following the cue onset. This is similar to the methodology used by @Thiery2016\: by measuring the timecourse of classifier accuracy, we are provided with an index of the deployment of spatial attention.

The second study extended this paradigm by adding a stimming condition. The accuracy of the classifier was compared between conditions where the participants were permitted to stim during the trials, or instructed to keep still. In this case, we used decoding accuracy to assess whether stimming provided a benefit in terms of the deployment of spatial attention; if there was a benefit, the classifier accuracy would be higher, as the EEG signal would differ more between conditions due to the greater focus of attention in different spatial locations. For the purposes of our analyses, due to an insufficiently sized and heterogeneous clinical group, we grouped the participants based on whether or not they reported stimming in their everyday lives; as discussed previously, neurotypicals often engage in stimming [@Chadehumbe2018], and 64% of our participants reported in the pre-experiment questionnaire that they typically engaged in stimming.

@@

41 participants took part in Experiment 1, but four produced very noisy data so were excluded, leaving a sample of 37 participants (20 male, 17 female). A total of 22 participants took part in Experiment 2 (8 male, 14 female). Participants in Experiment 2 were asked to report diagnoses of neurodevelopmental disorders – 17 had no diagnosis, 2 had a diagnosis of autism, and 3 had a diagnosis of ADHD. 15/22 participants further reported they typically engaged in stimming behaviours, while 7 reported that they did not.

@@

Stimuli were displayed on a ViewPixx 3D display (VPixx Technologies Inc., Quebec, Canada) with a diagonal extent of 24 inches, a refresh rate of 120Hz, and a resolution of 1920x1080 pixels. The screen was viewed by the participant from a distance of 57cm, resulting in a spatial resolution of 36 pixels per degree of visual angle. The display was gamma corrected using a Minolta LS110 photometer to ensure a linear luminance response. Stimuli were generated and displayed using a Mac Pro computer running Matlab, with the experiment script making use of functions from Psychtoolbox 3 [@Kleiner2007; @Brainard1997; @Pelli1997].

In both experiments, EEG recordings were acquired using an ANT Neuroscan 64-channel EEG system (ANT Neuro). Electrodes were mounted in a waveguard cap, arranged according to the 10/20 system, and were referenced to the whole head average with the ground electrode located at position *AFz*. EEG signals were recorded at 1kHz. In Experiment 2, participants’ eye movements were also recorded using an EyeLink 1000 (SR Research Ltd., Ottawa, Canada) eye tracker, also sampling at 1kHz. Low-latency digital triggers were sent from the stimulus computer to the EEG system using an 8-bit parallel cable.

@@

Raw EEG data were converted into EEGlab .set format using Matlab and the EEGlab toolbox [@Delorme2004]. We then used MNE python for the main analysis. Data were bandpass filtered between 0.2 and 30Hz, and then epoched using the event timestamps. We performed classification using a linear classifier with 10-fold cross-validation on the voltage data across all electrodes (excluding electrodes *M1* and *M2*), independently at each time point and for each participant. We then averaged classification accuracy across participants to generate the timecourses used in the main analysis. We calculate Bayes factors [@Rouder2009] to compare classification accuracy to chance performance (50\% correct) and to make between conditions and groups, but also report frequentist tests for comparison. We express Bayes factors in logarithmic units to account for extreme values ($log_{10}BF_{10}$), such that a Bayes factor of 3 is a $log_{10}BF_{10}$ = `r round(log10(3),digits=2)`, and constitutes convincing evidence for the alternative hypothesis, a Bayes factor of 10 is a $log_{10}BF_{10}$ = `r round(log10(10),digits=2)`, and constitutes strong evidence for the alternative hypothesis, and a Bayes factor of 30 is a $log_{10}BF_{10}$ = `r round(log10(30),digits=2)`, and constitutes overwhelming evidence for the alternative hypothesis.

@@

We first analysed the behavioural responses during the experiment, comparing accuracy and reaction times between stimuli shown in the cued (i.e. a congruent target) and uncued (an incongruent target) location. There was a highly significant effect of cue congruency on both accuracy ($log_{10}BF_{10}$ = `r round(log10(extractBF(ACCBF)$bf),digits=2)`, *t* = `r round(ACCtest$statistic,digits=2)`, *p* `r p_format(ACCtest$p.value,digits=3,accuracy=0.001)`, *d* = `r round(ACCd,digits=2)`) and reaction time ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTBF)$bf),digits=2)`, *t* = `r round(RTtest$statistic,digits=2)`, *p* `r p_format(RTtest$p.value,digits=3,accuracy=0.001)`, *d* = `r round(RTd,digits=2)`), as shown in @fig-Expt1resps. Responses to targets in the cued location were both faster ($M_{diff}$ = `r round(RTdiff)` ms) and more accurate ($M_{diff}$ = `r round(accdiff)`\%) than for those in the uncued location, showing a classic Posner cueing effect.

@@

In Experiment 2, we replicated the basic design of Experiment 1, but this time with an added 'stimming' condition, and with participants divided into groups based on their self-reported stimming behaviour. Pooling across all participants, for accuracy there was no effect of cue congruency ($log_{10}BF_{10}$ = `r round(log10(extractBF(accAOVBF)$bf[2]),digits=2)`, *F* = `r round(accAOVresults$ANOVA[2,4],digits=2)`, *p* = `r p_format(accAOVresults$ANOVA[2,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(accAOVresults$ANOVA[2,7],digits=2)`), stimming condition ($log_{10}BF_{10}$ = `r round(log10(extractBF(accAOVBF)$bf[1]),digits=2)`, *F* = `r round(accAOVresults$ANOVA[1,4],digits=2)`, *p* = `r p_format(accAOVresults$ANOVA[1,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(accAOVresults$ANOVA[1,7],digits=2)`), nor an interaction between the two ($log_{10}BF_{10}$ = `r round(log10(extractBF(accAOVBF)$bf[3]),digits=2)`, *F* = `r round(accAOVresults$ANOVA[3,4],digits=2)`, *p* = `r p_format(accAOVresults$ANOVA[3,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(accAOVresults$ANOVA[3,7],digits=2)`). For reaction time, there were significant effects of cue congruency ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBF)$bf[2]),digits=2)`, *F* = `r round(RTAOVresults$ANOVA[2,4],digits=2)`, *p* `r p_format(RTAOVresults$ANOVA[2,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresults$ANOVA[2,7],digits=2)`) and stimming condition ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBF)$bf[1]),digits=2)`, *F* = `r round(RTAOVresults$ANOVA[1,4],digits=2)`, *p* = `r p_format(RTAOVresults$ANOVA[1,5],digits=1,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresults$ANOVA[1,7],digits=2)`), as well as an interaction between the two ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBF)$bf[3]),digits=2)`, *F* = `r round(RTAOVresults$ANOVA[3,4],digits=2)`, *p* = `r p_format(RTAOVresults$ANOVA[3,5],digits=1,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresults$ANOVA[3,7],digits=2)`). Note that reactions were faster in the non-stimming condition (`r meanRTnostimming` ms) than in the stimming condition (`r meanRTstimming` ms). These effects remained significant in the group who reported stimming (N=`r sum(reportsstimming)`) for the effects of cue congruency ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBFSTIM)$bf[2]),digits=2)`, *F* = `r round(RTAOVresultsSTIM$ANOVA[2,4],digits=2)`, *p* < `r p_format(RTAOVresultsSTIM$ANOVA[2,5],digits=1,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresultsSTIM$ANOVA[2,7],digits=2)`), stimming condition ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBFSTIM)$bf[1]),digits=2)`, *F* = `r round(RTAOVresultsSTIM$ANOVA[1,4],digits=2)`, *p* = `r p_format(RTAOVresultsSTIM$ANOVA[1,5],digits=2,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresultsSTIM$ANOVA[1,7],digits=2)`), and their interaction ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBFSTIM)$bf[3]),digits=2)`, *F* = `r round(RTAOVresultsSTIM$ANOVA[3,4],digits=2)`, *p* = `r p_format(RTAOVresultsSTIM$ANOVA[3,5],digits=2,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresultsSTIM$ANOVA[3,7],digits=2)`). For the group who did not report stimming (N=`r length(reportsstimming)-sum(reportsstimming)`), only the effect of cue congruency was significant ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBFnoSTIM)$bf[2]),digits=2)`, *F* = `r round(RTAOVresultsNOSTIM$ANOVA[2,4],digits=2)`, *p* `r p_format(RTAOVresultsNOSTIM$ANOVA[2,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresultsNOSTIM$ANOVA[2,7],digits=2)`).

@@

The EEG results from Experiment 2 are summarised in @fig-Expt2EEG. Panels (a,b) demonstrate that multivariate pattern analysis is able to decode cue direction, target side, and target congruency with similar timecourses to Experiment 1 (see @fig-Expt1EEG), though with somewhat lower accuracy overall. We then compared decoding accuracy for cue side and cue congruency between the stimming group (N=14), and the non-stimming group (N=8). Although decoding accuracy appeared to be slightly higher for the no stimming group for cue side ([@fig-Expt2EEG]c), this difference was not convincing. Accuracy for cue congruency was very similar for the two groups ([@fig-Expt2EEG]d). Finally, we compared the deployment of spatial attention between stimming and non-stimming conditions for each of the two groups. There appeared to be a slight decrease in decoding accuracy in the non-stimming group when they were asked to stim ([@fig-Expt2EEG]f), but no differences in the stimming group ([@fig-Expt2EEG]e).

@@

Overall, our results from Experiment 1 support the assertion that EEG data can be used as a metric of attentional deployment [@Thiery2016]. The classifier was able to differentiate both cue and target side from EEG recordings in Experiment 1, indicating we have recorded the neural correlates of attentional deployment, and that it is possible for a machine learning classifier to differentiate which direction of attentional deployment these signals represent. Extending this method in Experiment 2 by including the stimming manipulation suggested that stimming does not have a detrimental effect on attentional deployment, with our only negative impact of stimming being on reaction times. We will now discuss these results in greater detail, with a focus on the implications of this for research into stimming in the context of the neurodiversity movement.

>>>>

if (processdata<2){
  # download processed data if missing
  if (!file.exists('local/Expt1resps.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Expt1resps.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
  if (!file.exists('local/Expt2resps.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Expt2resps.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
  if (!file.exists('local/Exp1Group.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Exp1Group.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
    if (!file.exists('local/Exp2Group.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Exp2Group.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
    if (!file.exists('local/Exp2GroupStim.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Exp2GroupStim.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}
    if (!file.exists('local/Exp2GroupNoStim.RData')){
osfproject <- osf_retrieve_node("dns89")
osffiles <- osf_ls_files(osfproject,n_max=300)
fid <- pmatch('Exp2GroupNoStim.RData',osffiles$name)
osf_download(osffiles[fid,],path='local/')}  
}

@@

  CIdecoding <- decodingdata
  CIdecoding$Cue <- apply(allCue,1,sd)/sqrt(ncol(allCue))*1.96
  CIdecoding$TargetU <- apply(allTargetU,1,sd)/sqrt(ncol(allTargetU))*1.96
  CIdecoding$CongIncong <- apply(allCongIncong,1,sd)/sqrt(ncol(allCongIncong))*1.96
  CIdecoding$TargetB <- apply(allTargetB,1,sd)/sqrt(ncol(allTargetB))*1.96
  CIdecoding$CueNostim <- apply(allCueNostim,1,sd)/sqrt(ncol(allCueNostim))*1.96
  CIdecoding$CueStim <- apply(allCueStim,1,sd)/sqrt(ncol(allCueStim))*1.96

@@

  CIdecoding <- decodingdata
  CIdecoding$Cue <- apply(allCue[,stimindices],1,sd)/sqrt(ncol(allCue[,stimindices]))*1.96
  CIdecoding$TargetU <- apply(allTargetU[,stimindices],1,sd)/sqrt(ncol(allTargetU[,stimindices]))*1.96
  CIdecoding$CongIncong <- apply(allCongIncong[,stimindices],1,sd)/sqrt(ncol(allCongIncong[,stimindices]))*1.96
  CIdecoding$TargetB <- apply(allTargetB[,stimindices],1,sd)/sqrt(ncol(allTargetB[,stimindices]))*1.96
  CIdecoding$CueNostim <- apply(allCueNostim[,stimindices],1,sd)/sqrt(ncol(allCueNostim[,stimindices]))*1.96
  CIdecoding$CueStim <- apply(allCueStim[,stimindices],1,sd)/sqrt(ncol(allCueStim[,stimindices]))*1.96

@@

        decodingBFs2 <- matrix(0,nrow=1201,ncol=7)
  for (t in 1:nrow(allCue)){
    decodingBFs2[t,1] <- extractBF(ttestBF(allCue[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,2] <- extractBF(ttestBF(allTargetU[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,3] <- extractBF(ttestBF(allCongIncong[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,4] <- extractBF(ttestBF(allTargetB[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,5] <- extractBF(ttestBF(allCueNostim[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,6] <- extractBF(ttestBF(allCueStim[t,nostimindices],mu=0.5,nullInterval=c(0.5,Inf)))$bf[1]
    decodingBFs2[t,7] <- extractBF(ttestBF(allCueStim[t,nostimindices],allCueNostim[t,nostimindices],paired=TRUE))$bf
  }
  CIdecoding <- decodingdata
  CIdecoding$Cue <- apply(allCue[,nostimindices],1,sd)/sqrt(ncol(allCue[,nostimindices]))*1.96
  CIdecoding$TargetU <- apply(allTargetU[,nostimindices],1,sd)/sqrt(ncol(allTargetU[,nostimindices]))*1.96
  CIdecoding$CongIncong <- apply(allCongIncong[,nostimindices],1,sd)/sqrt(ncol(allCongIncong[,nostimindices]))*1.96
  CIdecoding$TargetB <- apply(allTargetB[,nostimindices],1,sd)/sqrt(ncol(allTargetB[,nostimindices]))*1.96
  CIdecoding$CueNostim <- apply(allCueNostim[,nostimindices],1,sd)/sqrt(ncol(allCueNostim[,nostimindices]))*1.96
  CIdecoding$CueStim <- apply(allCueStim[,nostimindices],1,sd)/sqrt(ncol(allCueStim[,nostimindices]))*1.96

@@

A combined EEG and TMS study by @Capotosto2012 demonstrated that the application of TMS to the right intraparietal sulcus (IPS) impairs target detection, particularly on incongruent trials. They additionally noted that the P3 response amplitude was affected by this; on incongruent trials its amplitude was significantly lower, and on congruent trials it was significantly higher. The IPS has been previously identified as an important region in attentional deployment by @Vossel2009 in an fMRI study identifying regions that are selectively active on congruent and incongruent trials. The right IPS and, additionally, the right inferior frontal gyrus were implicated in attentional processes related to both types of trial. @Peelen2004 concluded that both endogenous and exogenous orienting cues involved the same network in their fMRI study, implicating frontal and parietal regions (the premotor cortex, posterior parietal cortex, medial frontal cortex, and right inferior frontal cortex). @Fitzgerald2015 conducted an MRI study using the Posner cueing task and found that activation in the ventral attention network differed between individuals with autism and controls; individuals with autism also showed weaker functional connectivity in the dorsal attentional network compared to their neurotypical counterparts.

@@

This study involved two experiments. In the first experiment, participants completed a Posner cueing task while EEG recordings were taken; a pattern classifier then attempted to differentiate, using the participants’ brain activity, whether they attended to the left vs. to the right following the cue onset. This is similar to the methodology used by @Thiery2016\: by measuring the timecourse of classifier accuracy, we are provided with an index of the deployment of spatial attention. The second experiment extended this paradigm by adding a stimming condition. The accuracy of the classifier was compared between conditions where the participants were permitted to stim during the trials, or instructed to keep still. In this case, we used decoding accuracy to assess whether stimming provided a benefit in terms of the deployment of spatial attention; if there was a benefit, the classifier accuracy would be higher, as the EEG signal would differ more between conditions due to the greater focus of attention in different spatial locations. For the purposes of our analyses, due to an insufficiently sized and heterogeneous clinical group, we grouped the participants based on whether or not they reported stimming in their everyday lives; as discussed previously, neurotypicals often engage in stimming [@Chadehumbe2018], and 64% of our participants reported in the pre-experiment questionnaire that they typically engaged in stimming.

@@

41 participants took part in Experiment 1, but four produced very noisy data so were excluded, leaving a sample of 37 participants (20 male, 17 female). A total of 22 participants took part in Experiment 2 (8 male, 14 female). Participants in Experiment 2 were asked to report diagnoses of neurodevelopmental disorders – 17 had no diagnosis, 2 had a diagnosis of autism, and 3 had a diagnosis of ADHD. 15/22 participants further reported they typically engaged in stimming behaviours, while 7 reported that they did not. Both experiments were granted ethical approval by the Ethics Committee of the Department of Psychology at the University of York.

@@

Stimuli were displayed on a ViewPixx 3D display (VPixx Technologies Inc., Quebec, Canada) with a diagonal extent of 24 inches, a refresh rate of 120Hz, and a resolution of 1920x1080 pixels. The screen was viewed by the participants from a distance of 57cm, resulting in a spatial resolution of 36 pixels per degree of visual angle. The display was gamma corrected using a Minolta LS110 photometer to ensure a linear luminance response. Stimuli were generated and displayed using a Mac Pro computer running Matlab, with the experiment script making use of functions from Psychtoolbox 3 [@Kleiner2007; @Brainard1997; @Pelli1997].

In both experiments, EEG recordings were acquired using an ANT Neuroscan 64-channel EEG system (ANT Neuro, Hengelo, Netherlands). Electrodes were mounted in a waveguard cap, arranged according to the 10/20 system, and were referenced to the whole head average. The ground electrode was located at position *AFz*, and EEG signals were recorded at 1kHz. In Experiment 2, participants’ eye movements were also recorded using an EyeLink 1000 (SR Research Ltd., Ottawa, Canada) eye tracker, also sampling at 1kHz. Low-latency digital triggers were sent from the stimulus computer to the EEG system using an 8-bit parallel cable.

@@

Raw EEG data were converted into EEGlab .set format using Matlab and the EEGlab toolbox [@Delorme2004]. We then used MNE python [@Gramfort2013] for the main analysis. Data were bandpass filtered between 0.2 and 30Hz, and then epoched using the event timestamps. We performed classification using a linear classifier with 10-fold cross-validation on the voltage data across all electrodes (excluding electrodes *M1* and *M2*), independently at each time point and for each participant. We then averaged classification accuracy across participants to generate the timecourses used in the main analysis. We calculate Bayes factors [@Rouder2009] to compare classification accuracy to chance performance (50\% correct) and to make comparisons between conditions and groups, but also report frequentist tests for reference. We express Bayes factors in logarithmic units to account for extreme values ($log_{10}BF_{10}$), such that a Bayes factor of 3 is a $log_{10}BF_{10}$ = `r round(log10(3),digits=2)`, and constitutes convincing evidence for the alternative hypothesis, a Bayes factor of 10 is a $log_{10}BF_{10}$ = `r round(log10(10),digits=2)`, and constitutes strong evidence for the alternative hypothesis, and a Bayes factor of 30 is a $log_{10}BF_{10}$ = `r round(log10(30),digits=2)`, and constitutes overwhelming evidence for the alternative hypothesis.

@@

We first analysed the behavioural responses from the experiment, comparing accuracy and reaction times between stimuli shown in the cued (i.e. a congruent target) and uncued (an incongruent target) location. There was a highly significant effect of cue congruency on both accuracy ($log_{10}BF_{10}$ = `r round(log10(extractBF(ACCBF)$bf),digits=2)`, *t* = `r round(ACCtest$statistic,digits=2)`, *p* `r p_format(ACCtest$p.value,digits=3,accuracy=0.001)`, *d* = `r round(ACCd,digits=2)`) and reaction time ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTBF)$bf),digits=2)`, *t* = `r round(RTtest$statistic,digits=2)`, *p* `r p_format(RTtest$p.value,digits=3,accuracy=0.001)`, *d* = `r round(RTd,digits=2)`), as shown in @fig-Expt1resps. Responses to targets in the cued location were both faster ($M_{diff}$ = `r round(RTdiff)` ms) and more accurate ($M_{diff}$ = `r round(accdiff)`\%) than for those in the uncued location, showing a classic Posner cueing effect.

@@

In Experiment 2, we replicated the basic design of Experiment 1, but this time with an added 'stimming' condition, and with participants divided into groups based on their self-reported stimming behaviour. Pooling across all participants, for response accuracy there was no effect of cue congruency ($log_{10}BF_{10}$ = `r round(log10(extractBF(accAOVBF)$bf[2]),digits=2)`, *F* = `r round(accAOVresults$ANOVA[2,4],digits=2)`, *p* = `r p_format(accAOVresults$ANOVA[2,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(accAOVresults$ANOVA[2,7],digits=2)`), stimming condition ($log_{10}BF_{10}$ = `r round(log10(extractBF(accAOVBF)$bf[1]),digits=2)`, *F* = `r round(accAOVresults$ANOVA[1,4],digits=2)`, *p* = `r p_format(accAOVresults$ANOVA[1,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(accAOVresults$ANOVA[1,7],digits=2)`), nor an interaction between the two ($log_{10}BF_{10}$ = `r round(log10(extractBF(accAOVBF)$bf[3]),digits=2)`, *F* = `r round(accAOVresults$ANOVA[3,4],digits=2)`, *p* = `r p_format(accAOVresults$ANOVA[3,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(accAOVresults$ANOVA[3,7],digits=2)`). For reaction time, there were significant effects of cue congruency ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBF)$bf[2]),digits=2)`, *F* = `r round(RTAOVresults$ANOVA[2,4],digits=2)`, *p* `r p_format(RTAOVresults$ANOVA[2,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresults$ANOVA[2,7],digits=2)`) and stimming condition ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBF)$bf[1]),digits=2)`, *F* = `r round(RTAOVresults$ANOVA[1,4],digits=2)`, *p* = `r p_format(RTAOVresults$ANOVA[1,5],digits=1,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresults$ANOVA[1,7],digits=2)`), as well as an interaction between the two ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBF)$bf[3]),digits=2)`, *F* = `r round(RTAOVresults$ANOVA[3,4],digits=2)`, *p* = `r p_format(RTAOVresults$ANOVA[3,5],digits=1,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresults$ANOVA[3,7],digits=2)`). Note that reactions were faster in the non-stimming condition (`r meanRTnostimming` ms) than in the stimming condition (`r meanRTstimming` ms). These effects remained significant in the group who reported stimming in their daily lives (N=`r sum(reportsstimming)`) for the effects of cue congruency ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBFSTIM)$bf[2]),digits=2)`, *F* = `r round(RTAOVresultsSTIM$ANOVA[2,4],digits=2)`, *p* < `r p_format(RTAOVresultsSTIM$ANOVA[2,5],digits=1,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresultsSTIM$ANOVA[2,7],digits=2)`), stimming condition ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBFSTIM)$bf[1]),digits=2)`, *F* = `r round(RTAOVresultsSTIM$ANOVA[1,4],digits=2)`, *p* = `r p_format(RTAOVresultsSTIM$ANOVA[1,5],digits=2,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresultsSTIM$ANOVA[1,7],digits=2)`), and their interaction ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBFSTIM)$bf[3]),digits=2)`, *F* = `r round(RTAOVresultsSTIM$ANOVA[3,4],digits=2)`, *p* = `r p_format(RTAOVresultsSTIM$ANOVA[3,5],digits=2,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresultsSTIM$ANOVA[3,7],digits=2)`). For the group who did not report stimming (N=`r length(reportsstimming)-sum(reportsstimming)`), only the effect of cue congruency was significant ($log_{10}BF_{10}$ = `r round(log10(extractBF(RTAOVBFnoSTIM)$bf[2]),digits=2)`, *F* = `r round(RTAOVresultsNOSTIM$ANOVA[2,4],digits=2)`, *p* `r p_format(RTAOVresultsNOSTIM$ANOVA[2,5],digits=3,accuracy=0.001)`, $\eta^2_G$ = `r round(RTAOVresultsNOSTIM$ANOVA[2,7],digits=2)`).

@@

The EEG results from Experiment 2 are summarised in @fig-Expt2EEG. Panels (a,b) demonstrate that multivariate pattern analysis is able to decode cue direction, target side, and target congruency with similar timecourses to Experiment 1 (see @fig-Expt1EEG), though with somewhat lower accuracy overall. We then compared decoding accuracy for cue side and cue congruency between the stimming group (N=14), and the non-stimming group (N=8). Although decoding accuracy appeared to be slightly higher for the non-stimming group for cue side ([@fig-Expt2EEG]c), this difference was not convincing. Accuracy for cue congruency was very similar for the two groups ([@fig-Expt2EEG]d). Finally, we compared the deployment of spatial attention between stimming and non-stimming conditions for each of the two groups. There appeared to be a slight decrease in decoding accuracy in the non-stimming group when they were asked to stim ([@fig-Expt2EEG]f), but no differences in the stimming group ([@fig-Expt2EEG]e).

@@

Overall, our results from Experiment 1 support the assertion that EEG data can be used as a metric of attentional deployment [@Thiery2016]. The classifier was able to differentiate both cue and target side from EEG recordings in Experiment 1, indicating we have recorded a neural correlate of attentional deployment, and that it is possible for a machine learning classifier to determine the direction of spatial attention using EEG signals. Extending this method in Experiment 2 by including the stimming manipulation suggested that stimming does not have a detrimental effect on attentional deployment, with our only negative impact of stimming being on reaction times. We will now discuss these results in greater detail, with a focus on the implications for research into stimming in the context of the neurodiversity movement.
